A Study on Common Bug Fix Patterns in Rust

### 2022-10-31 methodology.tex
methodology fixed

\todo{introductory para saying that we're going to describe our methodology}

@@

We now briefly describe our methodology, providing full details in later subsections. Our pipeline for finding classes of pervasive patterns in Rust programs parses program versions, computes differences between them, embeds them in fixed-size datapoints, and then clusters these points.

\todo[inline]{I think we only need to talk about this next bit in a subsection. It's an implementation detail, not fundamental to our approach: ``We decided to implement our pipeline completely in Python. To make this work, we need a way to get an AST as a Python dictionary.''}

\todo[inline]{I wonder if this should really go here, or if there should be a shorter version.}
To compute the contents of a change, we need two code revisions: the revision before the commit, and the revision after it. After parsing these two revisions, we will have two ASTs. Tree diff algorithms can compute the differences between two arbitrary trees. When the trees are programs' abstract syntax trees, we call their diff an ASTDiff. 
% , encoded in two Python dictionaries; We use dictdiffer to obtain ASTDiffs.
\todo[inline]{Maybe this next paragraph is what you want at the top of 2.1. Then in the next next paragraph you put a shorter description of the pipeline, in a sentence before ``we mined'', and you reference 2.1 there. Anyway think about organization and try to have something that makes sense.}
An ASTDiff may include an arbitrary number of semantic changes, although a best practice is to include only one semantic change in a commit. Because that best practice is not universally followed, we are interested in finding the most important change within an ASTDiff, and embedding that change in our datapoints. Sections~\ref{sec:path_extraction}~and~\ref{sec:weighting_scheme} provide a detailed explanation of how we select the most important semantic information, which then allowed us to obtain clusters that contained similar datapoints.

Our target repositories were the top 18 most-starred open source Rust projects on GitHub\todo{as of what date?}. We mined (using Pydriller) all the bug related commits and ran them through our pipeline; Section~\ref{sec:mining} describes our mining methodology in depth. Then, we applied the DBSCAN clustering algorithm~\cite{ester1996density} on the resulting datapoints. Section~\ref{sec:clustering_data} describes clustering in detail, justifies our choice of DBSCAN for clustering, and explains how we tuned it to improve cluster quality.

@@

\todo[inline]{should always be text between subsection and subsubsection. This is the per-change workflow, right?}

@@

Syn\footnote{\url{https://crates.io/crates/syn}} is a Rust crate built for the Rust procedural macro implementation. However, it includes a parser which is suitable for our purposes; we simply had to write a preprocessor to transform Syn AST output into Python dictionaries. Syn handles all of Rust. Each of our datapoints specifies whether any Syn non terminal (\texttt{NT}) is present or absent, i.e. containing one dimension per non terminal. However, we are also interested in change patterns that involve the borrow checker (BC), so we also add a dimension reflecting the presence of a set of BC-related elements (BCE) that we identified, e.g. \texttt{clone}, \texttt{Rc}, \texttt{Box}. The full list of \verb+BCE+ elements can be found in Appendix~\ref{sec:appendix}.

@@

The second part identifies the context of the modification, that is, the path from the root of the tree to the subtree in which the modification has occured. The third part is the content of the change---terminals and nonterminals. Our tool\todo{do we have a name? It's important to have a name} looks for changes within each root-level scope. That is, our assumption is that each change pattern occurs within one root-level scope. In this work, we chose to not handle patterns that involve changes in multiple functions or multiple files. However, our tool detects both continuous and non-continuous line changes within one scope.

Looking at the tree encoded in the content part of each diff, we realized that each path to a leaf represents one element contributing to the change. A Depth First Search allows us to collect all paths, and we return only the paths that we store in our datapoints (i.e. touching $\mathtt{NT} \cup \mathtt{BCE}$).
\todo{if you mathtt BCE here, you should do it everywhere, i.e. in the paragraph at the start of 2.1.1.}

@@

The most naive approach would be to only report the observed non terminals within the diff. That yields a fully order-insensitive representation. In such an embedding, for instance, there is no difference between two nested if statements and two if statements beside each other. On the other hand, theoretically, for full order-sensitiveness, we would need to embed all possible combinations of items from $\mathtt{NT} \cup \mathtt{BCE}$ as our dimensions, which yields an infeasible $O(n^d)$ dimensions ($n=|\mathtt{NT} \cup \mathtt{BCE}|$ and $d$ is the depth of nested structures); this could potentially be reduced somewhat, but is still impractical. While we can implement a specific workaround to distinguish nested ifs (and we have done so), the general point about the large dimensionality of fully order-sensitive embedding still holds.
\todo[inline]{Worse than that, isn't $d$ unbounded?}

@@

We propose a novel representation. Similar to the first\todo{which? the naive one?} approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCE}$ we collected from Syn. However, to simulate order-sensitiveness, we add two more steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension items \todo{is there an example?}. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs.

@@

% TODO fix this caption
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{clusters.pdf}
\caption{\label{fig:clustering} Clustering results with nine different combinations of parameters $\epsilon$ and $Z$, respectively separated in columns and rows. In each subplot, the two blue bars in the left respectively indicate the number of clustered ($G_c$) and noise ($G_n)$ datapoints (the number is shown above each bar) after applying clustering on general database $D_g$. The two orange bars in the right show similar data but for the borrow-checker $D_{b}$ database. In each subplot's legend, $C_G$ and $C_B$ show the number of clusters obtained from $D_g$ and $D_b$, respectively}
\end{figure}

\todo{figure caption: why the capital G and B in the subscript?}

@@

\todo{that is why what? huh? be more explicit and repeat yourself a bit}
That is why we carried out a manual analysis of clusters. In the manual analysis, we randomly pick 50 clusters; from each cluster we randomly choose 10 datapoints. The person analyzing the cluster \todo{who? be more specific about how we did this} then looks at the code and specifies the clusters that contained more than five datapoints showing similar patterns. For the manual analysis, we used $\epsilon=0.0001, Z=5$ and $\epsilon=0.001, Z=5$, which were the best parameters for $D_g$ and $D_b$, resulting in 577 and 102 clusters, respectively. As a final step, we exclude clusters that contained datapoints from fewer than 3 projects. That is because we wanted our clusters to contain cross-project (e.g. potentially generalizable) patterns as much as possible.

@@

\todo{can we say more-specific rather than concrete?}
In this work, we were only interested in the first two groups and ignored the clusters for refactoring changes. Also, we prefer reporting clusters that manifest concrete patterns. An abstract pattern as opposed to a concrete pattern can account for many changes in the programs. For instance, we deem the pattern `Changing a clone of a variable to a borrowing of it in a function argument' more concrete than `Adding a new statement to a function's body':

>>>>

% \todo{introductory para saying that we're going to describe our methodology}

In this section, we describe our methodology, providing full details in later subsections. Our pipeline for finding classes of pervasive patterns in Rust programs parses program versions, computes differences between them, embeds them in fixed-size datapoints, and then clusters these datapoints. After conducting a manual analysis on the obtained clusters, we refine them towards manifesting bug fix patterns.

@@

% \todo[inline]{I think we only need to talk about this next bit in a subsection. It's an implementation detail, not fundamental to our approach: ``We decided to implement our pipeline completely in Python. To make this work, we need a way to get an AST as a Python dictionary.''}

% \todo[inline]{I wonder if this should really go here, or if there should be a shorter version.}
% To compute the contents of a change, we need two code revisions: the revision before the commit, and the revision after it. After parsing these two revisions, we will have two ASTs. Tree diff algorithms can compute the differences between two arbitrary trees. When the trees are programs' abstract syntax trees, we call their diff an ASTDiff. 
% , encoded in two Python dictionaries; We use dictdiffer to obtain ASTDiffs.

We decided to implement our pipeline completely in Python. Our target repositories were the top 18 most-starred open source Rust projects on GitHub as of August 2021. We mined (using Pydriller) all the bug related commits and ran them through our pipeline; Section~\ref{sec:data_modelling} describes how we extracted fixed-size datapoints from the code changes. Section~\ref{sec:mining} describes our mining methodology in depth. Then, we applied the DBSCAN clustering algorithm~\cite{ester1996density} on the resulting datapoints. Section~\ref{sec:clustering_data} describes clustering in detail, justifies our choice of DBSCAN for clustering, and explains how we tuned it to improve cluster quality.

@@

% \todo[inline]{Maybe this next paragraph is what you want at the top of 2.1. Then in the next next paragraph you put a shorter description of the pipeline, in a sentence before ``we mined'', and you reference 2.1 there. Anyway think about organization and try to have something that makes sense.}
To compute the contents of a change, we need two code revisions: the revision before the commit, and the revision after it. After parsing these two revisions, we will have two ASTs. Tree diff algorithms can compute the differences between two arbitrary trees. When the trees are programs' abstract syntax trees, we call their diff an ASTDiff. An ASTDiff may include an arbitrary number of semantic changes, although a best practice is to include only one semantic change in a commit. Because that best practice is not universally followed, we are interested in finding the most important change within an ASTDiff, and embedding that change in our datapoints. Section~\ref{sec:parsing_programs} describes our program parsing process, and how we obtained ASTDiff from code changes. Sections~\ref{sec:path_extraction}~and~\ref{sec:weighting_scheme} provide a detailed explanation of how we select the most important semantic information out of an ASTDiff, which then allowed us to obtain clusters that contained similar datapoints.

% \todo[inline]{should always be text between subsection and subsubsection. This is the per-change workflow, right?}

@@

Syn\footnote{\url{https://crates.io/crates/syn}} is a Rust crate built for the Rust procedural macro implementation. However, it includes a parser which is suitable for our purposes; we simply had to write a preprocessor to transform Syn AST output into Python dictionaries. Syn handles all of Rust. Each of our datapoints specifies whether any Syn non terminal (\texttt{NT}) is present or absent, i.e. containing one dimension per non terminal. However, we are also interested in change patterns that involve the borrow checker (BC), so we also add a dimension reflecting the presence of a set of BC-related elements (\texttt{BCE}) that we identified, e.g. \texttt{clone}, \texttt{Rc}, \texttt{Box}. The full list of \texttt{BCE} elements can be found in Appendix~\ref{sec:appendix}.

@@

The second part identifies the context of the modification, that is, the path from the root of the tree to the subtree in which the modification has occured. The third part is the content of the change---terminals and nonterminals. Our tool\todo{do we have a name? It's important to have a name} looks for changes within each root-level scope (Rust book calls these root-level scopes the items of a crate\footnote{\url{https://doc.rust-lang.org/reference/items.html}}). That is, our assumption is that each change pattern occurs within one root-level scope. In this work, we chose to not handle patterns that involve changes in multiple functions or multiple files. However, our tool detects both continuous and non-continuous line changes within one scope.

Looking at the tree encoded in the content part of each diff, we realized that each path to a leaf represents one element contributing to the change. A Depth First Search allows us to collect all paths, and we return only the paths that we store in our datapoints (i.e. touching $\mathtt{NT} \cup \mathtt{BCE}$).
% \todo{if you mathtt BCE here, you should do it everywhere, i.e. in the paragraph at the start of 2.1.1.}

@@

The most naive approach would be to only report the observed non terminals within the diff. That yields a fully order-insensitive representation. In such an embedding, for instance, there is no difference between two nested if statements and two if statements beside each other. On the other hand, theoretically, for full order-sensitiveness, we would need to embed all possible combinations of items from $\mathtt{NT} \cup \mathtt{BCE}$ as our dimensions, which yields an infeasible $O(n^d)$ dimensions ($n=|\mathtt{NT} \cup \mathtt{BCE}|$ and $d$ is the depth of nested structures); this could potentially be reduced somewhat, but is still impractical. While we can implement a specific workaround to distinguish nested ifs (and we have done so), the general point about the large dimensionality of fully order-sensitive embedding still holds.
\todo[inline]{Worse than that, isn't $d$ unbounded? ``yes it is, but doesn't what we've said above still hold?''}

@@

We propose a novel representation. Similar to the naive approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCE}$ we collected from Syn. However, to simulate order-sensitiveness, we add two more steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension items\todo{is there an example? ``Fig 2?''}. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs.

@@

% TODO fix this caption
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{clusters.pdf}
\caption{\label{fig:clustering} Clustering results with nine different combinations of parameters $\epsilon$ and $Z$, respectively separated in columns and rows. In each subplot, the two blue bars in the left respectively indicate the number of clustered ($G_c$) and noise ($G_n)$ datapoints (the number is shown above each bar) after applying clustering on general database $D_g$. The two orange bars in the right show similar data but for the borrow-checker $D_{b}$ database. In each subplot's legend, $C_g$ and $C_b$ show the number of clusters obtained from $D_g$ and $D_b$, respectively}
\end{figure}

@@

Since both large and small values of $\epsilon$ would steer us away from obtaining meaningful clusters, we felt the need to carry out a manual analysis of clusters for better parameter tuning. In the manual analysis, we randomly pick 50 clusters; from each cluster we randomly choose 10 datapoints. The person analyzing the cluster (The first author) then looks at the code and specifies the clusters that contained more than five datapoints showing similar patterns. After conducting the manual analysis, we decided to use $\epsilon=0.0001, Z=5$ and $\epsilon=0.001, Z=5$, as our clustering parameters for $D_g$ and $D_b$, resulting in 577 and 102 clusters, respectively. As a final step, we exclude clusters that contained datapoints from fewer than 3 projects. That is because we wanted our clusters to contain cross-project (e.g. potentially generalizable) patterns as much as possible.

@@

In this work, we were only interested in the first two groups and ignored the clusters for refactoring changes. Also, we prefer reporting clusters that manifest more-specific patterns. An abstract pattern as opposed to a more-specific pattern can account for many changes in the programs. For instance, we deem the pattern `Changing a clone of a variable to a borrowing of it in a function argument' more specific than `Adding a new statement to a function's body':
